<div id="table-of-contents">
<h2>Table of Contents</h2>
<div id="text-table-of-contents">
<ul>
<li><a href="#sec-1">1. Grouping SNPs</a></li>
<li><a href="#sec-2">2. The groups and tags classes</a></li>
<li><a href="#sec-3">3. Automatic group building based on marginal posterior probabilities</a></li>
<li><a href="#sec-4">4. Manual corrections</a></li>
</ul>
</div>
</div>

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Groups, tags, and snp.picker}
-->
<font color="grey">
*Chris Wallace // [web](http://chr1swallace.github.io) // [email](mailto:cew54 at cam.ac.uk)*  
</font>

# Grouping SNPs<a id="sec-1" name="sec-1"></a>

We often (usually) cannot resolve an association to a single causal
variant, and one of the aims of GUESSFM is to generate groups of SNPs,
one of which is likely to be causal.

# The groups and tags classes<a id="sec-2" name="sec-2"></a>

The `groups` and `tags` classes are two ways to represent the same
information, but use different structures which might be more
convenient for one function or another.

```{r }
library(GUESSFM)
showClass("groups")
showClass("tags")
```

# Automatic group building based on marginal posterior probabilities<a id="sec-3" name="sec-3"></a>

Once models have been fitted, we can average over them to find
marginal posterior probabilities for each SNP.  We make the assumption
that SNPs in a group should display the following properties:

-   the sum of MPI should be no more than 1
-   SNPs in a group should not occur in the same model together very
    often
-   SNPs in a group should be correlated (r<sup>2</sup>>0.5), and the MPI should tail
    off from the primary SNP with r<sup>2</sup>

`snp.picker()` aims to create such groups in an automated manner,
given the summary of a GUESS run and a genotype dataset from which r<sup>2</sup>
may be calculated.  It returns an object of class `snppicker` which is
an extension of the `groups` class and contains information to allow
plotting of the relationship between r<sup>2</sup> and MPI that was used to
define a group.

```{r }
showClass("snppicker")
```

# Manual corrections<a id="sec-4" name="sec-4"></a>

Plotting an object of class `snppicker` should be informative.  The
algorithm underlying `snp.picker()` has been designed so it is more
likely to err by splitting a single group in two than by joining what
might be two separate groups.  If you suspect two groups should be
merged, you can check the assumptions above using `check.merge()`, and
perform a merge using `group.merge()`.
